---
title: "Data"
author: "Michael McDonagh (2392623)"
date: "2026-10-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Data 

# Import

First we automatically import the data for 23 seasons of English football from the top 4 divisions.
We are pulling all our data from Football-Data.co.uk, a site containing results, match statistics and betting odds since the 2003/2004 season, and basic stats since the second Premier League season in 1993/94.

```{r}

if(!dir.exists(file.path("..", "data"))){
  dir.create(file.path("..", "data"))
}

years<-2003:2025

# Splitting up the leagues

leagues<-c(
  E0 = "Premier_League",
  E1 = "Championship",
  E2 = "League_1",
  E3 = "League_2"
)

for(year in years){
  season<-paste0(
    substr(year, 3, 4),
    sprintf("%02d", (year+1) %% 100)
  )
  
  for(league_code in names(leagues)){
    league_name<-leagues[[league_code]]
    
    
    file_name<-paste0(league_name, "_data_", season, ".csv")
    file_path<-file.path("..", "data", file_name)
    
# Only downloading if not already there
    
    if(!file.exists(file_path)){
      url<-paste0("https://football-data.co.uk/mmz4281/", season, "/", league_code, ".csv")
      download.file(url, file_path)
    }
  }

}
```

Next creating a compiled data frame: all 4 leagues across all 23 seasons.

```{r}
files<-list.files(
  file.path("..", "data"),
  pattern = "\\.csv$",
  full.names = TRUE
)

data_list<-list()

for(i in seq_along(files)){
  df<-read.csv(files[i])
  
# This deals with a bug in the data. Some rows of certain seasons had empty strings,
# which caused additional matches to be added with no information.
  df<-df[!is.na(df$Date) & df$Date!="", ]
  
  file_name<-basename(files[i])
  
# This creates a Season variable, by recording it from the file name
  df$Season<-substr(
    file_name,
    nchar(file_name)-7,
    nchar(file_name)-4
  )
  
# Likewise this creates a League variable
  if(grepl("Premier_League", file_name)){
    df$League<-"Premier League"
  }
  else if(grepl("Championship", file_name)){
    df$League<-"Championship"
  }
  else if(grepl("League_1", file_name)){
    df$League<-"League 1"
  }
  else if(grepl("League_2", file_name)){
    df$League<-"League 2"
  }
  
  data_list[[i]]<-df
}

# This ensures that the dimensions of all the data are the same by creating N/A columns for variables
# that aren't recorded in certain seasons 
all_columns<-unique(unlist(lapply(data_list, names)))
data_list<-lapply(data_list, function(df){
  missing_columns<-setdiff(all_columns, names(df))
  df[missing_columns]<-NA
  df<-df[all_columns]
  return(df)
})

# Creating the data frame
data<-do.call(rbind, data_list)
```

```{r}
# Export to send for use in Python

write.csv(
  data,
  "../data/football_data_halfclean.csv",
  row.names = FALSE
)
```

# Exploratory Data Analysis

We will create some important visualistation of the data to get a broader picture of it.
Firstly, a table showing how the outcomes H (Home Win), D (Draw), and A (Away Win) are distributed for each 
league over the 23 seasons:

```{r}
prop.table(table(data$FTR, data$League), margin = 2)
```





